Measuring contrast processing in the visual system using SSVEP

### 2024-04-13 review.qmd
Minor edits

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@McKeefry1996; @Baseler1997; @Di-Russo2001].

Contrast is relatively simple to define: typically, contrast is specified as the percentage deviation of a uniform stimulus from the background. So, for example, a disk of 100 units of cone activation ($I_{\mathrm{stim}}$) surrounded by a 'background' of 50 units of activation ($I_{\mathrm{background}}$) has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the @Michelson1927 definition of contrast is specified by the maximum and minimum excursions from the mean:

@@

```{python}
#| label: fig-centresurround
#| fig-cap: The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region"
#| fig-cap-location: bottom

@@

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions reported by studies measuring single unit activity or local field potentials in the cortex [@Shapley1980; @Morrone1982]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

@@

Previous figures showed how a single-input SSVEP might generate a single set of peaks in the frequency domain response which, in turn, can be used as a read-out of neuronal activity. In @fig-plotspectra, we show the effect of adding a second contrast component: the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988]

@@

```{python}
#| label: fig-plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom

@@

```{python}
#| label: fig-plotspectra
#| fig-cap: "Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz combination. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2"
#| fig-cap-location: bottom

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

```{python}
#| label: fig-contrastresponsefunction
#| fig-cap: "Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels."
#| fig-cap-location: bottom

@@

```{python}
#| label: fig-intermodulation
#| fig-cap: "Non-linear combination of two independent contrast modulations at F1 and F2 results in 'intermodulation' terms that appear at low-order sums and differences of those frequencies"
#| fig-cap-location: bottom

@@

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex contains is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal.

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. @Candy2001; @Busse2009; @Tsai2012; @Baker2017 *** SEE ALSO RECENT WORK FROM WINAWER LAB - MUCH OF IT IS 'FAST' VEPs ***

@@

#| label: fig-sweepvep
#| fig-cap: "Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts"
#| fig-cap-location: bottom

@@

Parkinson's

Alex - it would be good to add some CRFs from @Afsari2014 to the figure. Is it possible to extract either the fitted parameter values or the raw data for some control flies and some mutants? Whatever you think is most compelling.

```{python}
#| label: fig-plotclinical
#| fig-cap: "Example contrast response functions for different clinical conditions. Panel (a) shows modelled contrast response functions for epilepsy patients (blue) and control participants (black), based on the data of Tsai et al. (2011). Panel (b) shows functions for the amblyopic (red) and fellow (black) eyes of adults with amblyopia, based on the data of Baker et al. (2015). Panel (c) shows functions for children with (green) and without (black) a diagnosis of autism, based on the data of Vilidaite et al. (2018). Panel (d) shows data from *Drosophila melanogaster* (fruit flies) in Afsari et al 2014. One day-old flies expressing a human gene linked to Parkinson's (hLRRK-G2019S) show increased SSVEP response amplitude and sensitivity compared to control animals"
#| fig-cap-location: bottom

>>>>

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@McKeefry1996; @Baseler1997; @Di-Russo2001].

Contrast is relatively simple to define: typically, it is specified as the percentage deviation of a uniform stimulus from the background. So, for example, a disk of 100 units of cone activation ($I_{\mathrm{stim}}$) surrounded by a 'background' of 50 units of activation ($I_{\mathrm{background}}$) has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the @Michelson1927 definition of contrast is specified by the maximum and minimum excursions from the mean:

@@

```{python}
#| label: fig-centresurround
#| fig-cap: The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region
#| fig-cap-location: bottom

@@

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency, or a low multiple thereof. This corresponds closely to similar functions reported by studies measuring single unit activity or local field potentials in the cortex [@Shapley1980; @Morrone1982]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

@@

**Previous figures** showed how a single-input SSVEP might generate a single set of peaks in the frequency domain response which, in turn, can be used as a read-out of neuronal activity. In @fig-plotspectra, we show the effect of adding a second contrast component: the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988]

@@

```{python}
#| label: fig-plotsinewaves
#| fig-cap: a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain.
#| fig-cap-location: bottom

@@

```{python}
#| label: fig-plotspectra
#| fig-cap: Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2.
#| fig-cap-location: bottom

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

```{python}
#| label: fig-contrastresponsefunction
#| fig-cap: Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels.
#| fig-cap-location: bottom

@@

```{python}
#| label: fig-intermodulation
#| fig-cap: Non-linear combination of two independent contrast modulations at F1 and F2 results in 'intermodulation' terms that appear at low-order sums and differences of those frequencies
#| fig-cap-location: bottom

@@

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal.

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. @Candy2001; @Busse2009; @Tsai2012; @Baker2017 *** SEE ALSO RECENT WORK FROM WINAWER LAB - MUCH OF IT IS 'FAST' VEPs ***

Adaptation

@@

#| label: fig-sweepvep
#| fig-cap: Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts.
#| fig-cap-location: bottom

@@

Parkinson's @Afsari2014

```{python}
#| label: fig-plotclinical
#| fig-cap: Example contrast response functions for different clinical conditions. Panel (a) shows modelled contrast response functions for epilepsy patients (blue) and control participants (black), based on the data of Tsai et al. (2011). Panel (b) shows functions for the amblyopic (red) and fellow (black) eyes of adults with amblyopia, based on the data of Baker et al. (2015). Panel (c) shows functions for children with (green) and without (black) a diagnosis of autism, based on the data of Vilidaite et al. (2018). Panel (d) shows data from *Drosophila melanogaster* (fruit flies) in Afsari et al 2014. One day-old flies expressing a human gene linked to Parkinson's (hLRRK-G2019S) show increased SSVEP response amplitude and sensitivity compared to control animals.
#| fig-cap-location: bottom
